---
title: "Reproducible Research: Peer Assessment 1"
author: "TZ"
date: "Thursday, January 18, 2015"
output: 
  html_document:
    keep_md: true
---


## Introduction
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  
The data for this assignment can be downloaded from the course web
site:

Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

 **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

 **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

 **interval**: Identifier for the 5-minute interval in which
    measurement was taken


The dataset is stored in a comma-separated-value (CSV) file and there
are a total of 17,568 observations in this
dataset.


## Loading and preprocessing the data

If csv-file "activity.csv" with needed data set does not exist in working directory following fragment of code will download (if it is necessary) the zip-file into working_directory and unzip its.

```{r filedownloading}
if( !file.exists("activity.csv") ) {
        if ( !file.exists("repdata_data_activity.zip") ) {
                ## downloading and unzipping file into 'working_directory/data'
                library(downloader)
                url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
                download(url, "repdata_data_activity.zip")
                unlink(url)
                rm(url) ## removing spent variable
        }
        unzip("repdata_data_activity.zip", exdir = "./data")
}
```
Then is reading file into data frame named 'activity'
```{r filereading}
activity <- read.csv("activity.csv",
                     sep = ",",
                     header = TRUE,
                     na.strings="NA", 
                     stringsAsFactors = FALSE)
```

Exploration of data frame:
```{r dataExploration}
head(activity)
tail(activity)
str(activity)
summary(activity)
dates <- unique(activity$date); # dates in character format
length(dates);
intervals <- unique(activity$interval); 
length(intervals); 
print(intervals)
```
You can see that the **sequence number of 5-minut interval is broken**.
24 hours x (60 min/hours / (5 min/observation)) = 288. **It is ok.**  
24 hours x 60 min/hours = **1440 min**.  
But in data set 'activity' we have intervals **from 0 to 2355 minutes**. That is **not** ok!

```{r }
nIntervals <- 24*(60/5)
print(nIntervals)
```

Now I process and transform the data into a format suitable for following analysis:
I perform column 'date' from 'character' to 'Date' format, add column 'numOfInterval' with sequence number of 5-minut interval, and  correct column 'interval'
```{r transforming0}
activity = transform( activity, date = as.Date(date), 
                     numOfInterval = rep( c(1:nIntervals), length(dates) ), 
                     interval = rep( ((seq(1:nIntervals)*5)-5), length(dates)))
summary(activity)
dates <- unique(activity$date); # dates in Date format
intervals <- unique(activity$interval); # true intervals
```


## What is mean total number of steps taken per day?

```{r }
total_steps_each_day <- tapply(activity$steps, 
                              activity$date, 
                              FUN = sum, 
                              simplify = TRUE)
summary(total_steps_each_day)
```

```{r plot1}
par(mfrow = c(1,1), mar=c(6,4,1,1))
plot(dates, total_steps_each_day, 
#      type = "l",
     xlab = "Dates",
     ylab = "The total number of steps per day")
lines(dates, total_steps_each_day)
```


Number of missing values (NAs) in vector 'total_steps_each_day' is equel:
```{r }
sum(is.na(total_steps_each_day))
sum(is.na(total_steps_each_day))/length(dates)*100 # in percent
```

The total number of steps taken each day is NA in following days:
```{r }
dates[is.na(total_steps_each_day)]
```
What does it mean?  
I don't see in these sequences of dates any pattern. Maybe some days an anonymous forgot about using of the device for counting taken steps. 


Histogram of the total number of steps taken each day:
```{r plot2}
par(mfrow = c(2,1), mar=c(4,4,4,1))
hist(total_steps_each_day, 
     breaks = 10,
     col = "cyan",
     main = "Histogram of the total number of steps\n taken each day",
     font.main = 12,
     xlab = "")
text(20000, 8,"breaks = 10")
hist(total_steps_each_day, 
     breaks = 20,
     col = "magenta",
     xlab = "Total number of steps taken each day",
     main = "")
text(20000, 8,"breaks = 20")
```

Mean and median of total number of steps taken per day are
```{r }
mean(total_steps_each_day, na.rm = TRUE)
median(total_steps_each_day, na.rm = TRUE)
```


## What is the average daily activity pattern?
```{r }
avr_steps_each_interval <- tapply(activity$steps, 
                               activity$interval, 
                               FUN = mean, 
                               na.rm = TRUE, 
                               simplify = TRUE)
```

```{r plot3}
par(mfrow = c(1,1), mar=c(4,4,1,1))
plot(intervals, avr_steps_each_interval, 
     type = "l",
     xlab = "Interval",
     ylab = "Average number of steps taken, averaged across all days")
```

Ordinal number of 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps is equel
```{r }
ind_avr <- which.max(avr_steps_each_interval)[[1]]
print(ind_avr)
```

In is following 5-minute interval:
```{r }
(ind_avr-1)*5
```


What is it in hours?
```{r }
(ind_avr-1)*5/60
```


Maximum of average number of steps in interval is
```{r }
avr_steps_each_interval[[ind_avr]]
```



## Imputing missing values

Only column 'steps' contains missing values (NA).  
I calculate now the total number of missing values
```{r }
rows_with_NA <- is.na(activity$steps)
nrows_with_NA <- sum(rows_with_NA)
print(nrows_with_NA)
nrows_with_NA/nrow(activity)*100 # in percent
```
It is not very much but the presence of missing days may introduce bias into some calculations or summaries of the data.  
What is strategy for filling in all of the missing values in the dataset?
"The strategy does not need to be sophisticated".
I will try to use 2 strategies:  
1. NAs replacing by the **median** for that 5-minute interval  
2. NAs replacing by the **mean** for that 5-minute interval  

```{r }
activity -> activity1 -> activity2
ind <- (activity1$numOfInterval[rows_with_NA])
```

Filling in all of the missing values in the dataset 'activity1' with the **median** for that 5-minute interval
```{r }
med_steps_each_interval <- tapply(activity$steps, 
                                  activity$interval, 
                                  FUN = median, 
                                  na.rm = TRUE, simplify = TRUE)
```
```{r activity1}
activity1$steps[rows_with_NA] <- med_steps_each_interval[ind]
```

Filling in all of the missing values in the dataset 'activity2' with the **mean** for that 5-minute interval
```{r activity2}
activity2$steps[rows_with_NA] <- avr_steps_each_interval[ind]
```

```{r }
total1_steps_each_day <- tapply(activity1$steps, 
                               activity1$date, 
                               FUN = sum, 
                               na.rm = TRUE, simplify = TRUE)

total2_steps_each_day <- tapply(activity2$steps, 
                                activity2$date, 
                                FUN = sum, 
                                na.rm = TRUE, simplify = TRUE)
```

Histogram of the total number of steps taken each day
```{r plot4}
par(mfrow = c(2,1), mar=c(4,4,4,1))
hist(total1_steps_each_day, 
     breaks = 10,
     col = "blue",
     ylim = c(0, 25),
     main = "Histogram of the total number of steps\n taken each day",
     font.main = 12,
     xlab = "")
text(17000, 20,"NAs are replaced by median")

hist(total2_steps_each_day, 
     breaks = 10,
     col = "red",
     ylim = c(0, 25),
     xlab = "Total number of steps taken each day",
     main = "")
text(17000, 20,"NAs are replaced by mean")
```

After using 2nd strategy (NAs are replaced by mean) data set better describes real life activity of subject. I think that subject in his/her life hasn't got 10 days from 2 month without moving when total number of steps per day is less than 2000).

Mean and median of total number of steps taken per day after realizing *1st* strategy:
```{r }
mean(total1_steps_each_day)
median(total1_steps_each_day)
```

These values differ from mean (particular) and median of total number of steps taken per day in data set *with NAs*.
 
 
Mean and median of total number of steps taken per day after realizing *2nd* strategy:
```{r }
mean(total2_steps_each_day)
median(total2_steps_each_day)
```

These values are practically equal mean and median of total number of steps taken per day in data set *with NAs*. That is very well.


```{r }
avr1_steps_each_interval <- tapply(activity1$steps, 
                                   activity1$interval, 
                                   FUN = mean, 
                                  na.rm = TRUE, simplify = TRUE)
avr2_steps_each_interval <- tapply(activity2$steps, 
                                   activity2$interval, 
                                   FUN = mean, 
                                   na.rm = TRUE, simplify = TRUE)
```

```{r plot5} 
par(mfrow = c(1,1), mar=c(4,4,1,1))
plot(intervals, avr2_steps_each_interval, 
     type = "l", col = "red",
     xlab = "Interval", 
     ylab = "Average number of steps taken, averaged across all days")
lines(intervals, avr1_steps_each_interval, 
     col = "blue")
legend(x=600, y=200, lty = "solid", 
       col = c("blue", "red"), 
       legend = c("NAs are replaced by median", "NAs are replaced by mean"),
       bty = "n")
```                                   

My conclusion, 2nd strategy (NAs are replaced by mean) in this case is better than 1st strategy (NAs are replaced by median). I will data set 'activity2' for next part of this assignment.

## Are there differences in activity patterns between weekdays and weekends?

Creatinga new data frame 'activity3' on basis of 'activity2' and new variables in this dataset:  
-character vector with names of weekdays ('week.days')  
-and factor 'wd' with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r }
activity3 = transform(activity2, 
                      week.days = weekdays(date, abbreviate = TRUE),
                      wd = as.character("weekday"),
                      stringsAsFactors = FALSE) # !!!!!!
wkendIndex <- (activity3$week.days == "Sun" | activity3$week.days == "Sat")
activity3$wd[wkendIndex] <- "weekend"
activity3 <- transform(activity3, wd = factor(wd))
```

Creating two separated data sets for weekends and weekdays using logical vector 'wkendIndex':
```{r }
activity3wkday <- activity3[!wkendIndex,]
activity3wkend <- activity3[wkendIndex,]
```

Calculation the average number of steps taken per day, averaged across all weekday days or weekend days:
```{r }
avr3_steps_each_interval_wkday <- tapply(activity3wkday$steps, 
                                         activity3wkday$interval,
                                         FUN = mean, 
                                         na.rm = TRUE, simplify = TRUE)
avr3_steps_each_interval_wkend <- tapply(activity3wkend$steps, 
                                         activity3wkend$interval,
                                         FUN = mean, 
                                         na.rm = TRUE, simplify = TRUE)
```

Making a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r plot6}
ymax_on_plot6 <- round(max(avr3_steps_each_interval_wkend, avr3_steps_each_interval_wkday))+10
par(mfrow = c(2,1), mar=c(4,4,1.5,1))
plot(intervals, avr3_steps_each_interval_wkend, 
     type = "l", 
     xlim = c(0, nIntervals*5), 
     ylim = c(0, ymax_on_plot6), 
     xlab = "", ylab = "Number of steps",
     main = "weekend", 
     font.main = 12, 
     font.lab = 10,  
     font.axis = 8,
     font = 1)
plot(intervals, avr3_steps_each_interval_wkday, 
     type = "l", 
     xlim = c(0, nIntervals*5), 
     ylim = c(0, ymax_on_plot6), 
     xlab = "interval", ylab = "Number of steps",
     main = "weekday",
     font.main = 12, 
     font.lab = 10,  
     font.axis = 8,
     font = 1)
```


###Answer:  
Yes, there are differences in activity patterns between weekdays and weekends.
On weekends subject have more of going/running than on weekdays. Perhups subject have sedentary job.